Compilation and Backend-Independent Vectorization for Multi-Party Computation

### 2022-04-30 overleaf/vectorization.tex
Edits to Section 5.

\ana{Need an intro/overview here!}
\ana{Need to define notion of canonical access!}
\ana{Talk about MPC Source, opportunities for vectorization, algorithm unique to MPC Source.}

@@

The next steps construct the dependence graph and \ana{Add an intro here.}

\subsubsection{Def-use Edges}

The dependence graph has the following def-use edges:

@@

\subsubsection{Closures}

We define $\mathit{closure}(n)$ where $n$ is a phi-node. Intuitively, it computes the set of nodes (i.e., statements) that form a dependence cycle with $n$. The closure of $n$ is defined as follows:
\begin{itemize}
\item $n$ is in $\mathit{closure}(n)$
\item $X$ is in $\mathit{closure}(n)$ if there is a same-level path from $n$ to $X$, and $X \rightarrow n$ is a same-level back-edge.
\item $Y$ is in $\mathit{closure}(n)$ if there is a same-level path from $n$ to $Y$ and there is a same-level path from $Y$ to some $X$ in $\mathit{closure}(n)$.
\end{itemize}

\subsection{Scalar Expansion}

@@

An important component of our algorithm is the scalar expansion to the corresponding loop dimensionality, 
which is necessary to expose opportunities for vectorization.
For example, {\sf d = S[i*D+j] - C[j]} equiv. to {\sf d = S[i,j] - C[j]}, which gave rise to {\sf N*D} subtraction operations in the sequential schedule,
is lifted. The argument arrays {\sf S} and {\sf C} are lifted and the scaler {\sf d} is lifted: {\sf d[i,j] = S[i,j] - C[i,j]}.
The algorithm then detects that the statement can be vectorized.

\paragraph{Arrays} There are three kinds of arrays.
\begin{itemize}

\item Scalars: These are scalar variables we expand into arrays for the purposes of vectorization.
For those, all writes are canonical writes and all reads are canonical reads. We will \emph{raise dimension} when
a scalar gives rise to an outer-to-inner dependence edge (e.g., {\sf sum!2} in line 6 of the MPC source code will be raised to a 1-dimensional
array since {\sf sum!2} is used in the inner $j$-loop). We will \emph{drop dimension} when a scalar
gives rise to an inner-to-outer dependence edge (e.g., {\sf sum!2} for which the lifted inner loop computes
{\sf D} values, but the outer loop only needs the last one.)

@@

In the sections below we detail the $\mathit{raise\_dim}$ (raise dimension) and $\mathit{drop\_dim}$ (drop dimension) operations,
followed by our vectorization algorithm.

@@

\paragraph{Raise dimension} 
There are two conceptual versions of $\mathit{raise\_dim}$. One reshapes an arbitrary access into a canonical read access in the corresponding loop. 
%The signature of $\mathit{raise\_dim}$ is as follows. 
It takes the original array, the access pattern function $f(i,j,k)$ in loop nest $i,j,k$ and the loop bounds
$((i:I),(j:J),(k:K))$:
\[ \mathit{raise\_dim}({\sf A, f(i,j,k), ((i:I),(j:J),(k:K))}) \]
It produces a new 3-dimensional array {\sf A'} by iterating over $i,j,k$ and setting each element of  {\sf A'} as follows:
\[ {\sf A'[i,j,k] =  A[f(i,j,k)]} \]
The end result is that uses of {\sf [A[f(i,j,k)]} in loop nest $i,j,k$ are replaced with canonical read-accesses to {\sf A'[i,j,k]}
that can be vectorized. In the running Biometric example, {\sf C'} = $\mathit{raise\_dim}${\sf (C, j, (i:N,j:D))} lifts the
1-dimensional array {\sf C} into a 2-dimensional array. The $i,j$ loop now accesses {\sf C'} in the canonical way, {\sf C'[i,j]}.
Similarly, {\sf S'} = $\mathit{raise\_dim}${\sf (S, i*D+j, (i:N,j:D))} tries to lift {\sf S}, but the operation turns into a no-op
because {\sf S} is already a 2-dimensional array and the read access is canonical.

\paragraph{Drop dimension} 
The other version of $\mathit{raise\_dim}$ lifts a lower-dimension array into a
higher-dimension for access in a nested loop. It is necessary when processing outer-to-inner dependences. 
Here {\sf A} is an $i$-array and raise dimension adds two additional dimensions:
\[ \mathit{raise\_dim}({\sf A}, {\sf (j:J,k:K)}) \]
This version is reduced to the above version by adding the access pattern function, which is just $i$:
\[ \mathit{raise\_dim}({\sf A, i, (j:J,k:K)}) \]

The corresponding $\mathit{drop\_dim}$ is carried out when an array written in an inner loop is used in an enclosing loop.
It takes a higher dimensional array, say $i,j,k$ and removes trailing dimensions, say $j,k$:
\[ \mathit{drop\_dim}({\sf A, (j:J,k:K)}) \]
It iterates over $i$ and takes the result at the maximal index of $j$ and $k$, i.e., the result at the last iterations of $j$ and $k$:
\[ {\sf A'[i] = A[i,J-1,K-1]} \]

\subsection{Basic Vectorization}

We present our vectorization algorithm in detail followed by a detailed example on Biometric Matching.

@@

\FOR {each MPC $\mathit{stmt}: X = \mathit{Op}(Y_1,Y_2)$ in loop $i,j,k$}
%\FOR {each $Y_n$}
\FOR {each argument $Y_n$} %def-use edge $\mathit{stmt}' (\mbox{def of } Y) \rightarrow \mathit{stmt} (\mbox{def of } X)$}
\STATE {case def-use edge $\mathit{stmt}' (\mbox{def of } Y_n) \rightarrow \mathit{stmt} (\mbox{def of } X)$ of} \\
\STATE \hspace{0.25cm} {same-level: $Y'_n$ is $Y_n$} \\
\STATE \hspace{0.25cm} {outer-to-inner: add $Y'_n[i,j,k] = \mathit{raise\_dim}(Y_n)$ at $\mathit{stmt'}$}
\STATE \hspace{0.25cm} {(more precisely, right after $\mathit{stmt'}$)}
\STATE \hspace{0.25cm} {inner-to-outer: add $Y'_n[i,j,k] = \mathit{drop\_dim}(Y_n)$ at $\mathit{stmt}$}
\STATE \hspace{0.25cm} {(more precisely, in loop of $\mathit{stmt}$ right after loop of $\mathit{stmt'}$)}
\ENDFOR

\COMMENT { Optimistically vectorize all. $I$ means vectorized dimension. }
%\FOR {each MPC $\mathit{stmt}: X = \mathit{Op}(Y_1,Y_2)$ in loop $i,j,k$}
\STATE {change to $X[I,J,K] = \mathit{Op}(Y'_1[I,J,K],Y'_2[I,J,K])$ }
\ENDFOR

\COMMENT { Phase 2: Recreating FOR loops for cycles; vectorizable statements hoisted up. }

\FOR {each dimension $d$ from highest to 0}
\FOR {each $\phi$-node $n$ in loop $i_1,...,i_d$}
\STATE {compute $\mathit{closure}(n)$}
\ENDFOR

@@

\WHILE {there are closure $cl_1$ and $cl_2$ that intersect}
\STATE {merge $cl_1$ and $cl_2$}
\ENDWHILE
\FOR {each closure $cl$ (after merge)}
\STATE {create FOR $i_d = 0; ...$ loop}
\STATE {add $\phi$-nodes in $cl$ to header block}
\STATE {{\color{red}{add target-less $\phi$-node for A if $cl$ updates array A}}}
\STATE {add statements in $cl$ to loop body in some order of dependences}
\STATE \COMMENT { Dimension is not vectorizable: }
\STATE {change ${I_d}$ to $i_d$ in all statements in loop}
\STATE {treat FOR loop as monolith node for def-uses: e.g., some def-use edges become same-level.}
\ENDFOR
{
\color{red}
\FOR {each target-less $\phi$-node $A_1 = \phi(A_0,A_k)$}
\STATE {in vectorizable stmts, replace use of $A_1$ with $A_0$}
\STATE {discard $\phi$-node if not used in any $cl$, replacing $A_1$ with $A_0$ or $A_k$ appropriately}
\ENDFOR
}
\ENDFOR

@@

\COMMENT { A dimension $i$ is dead on exit from stmt $X[...i...] = ...$ if all def-uses with targets outside of the enclosing
FOR $i=0 ...$ MOTION loop end at target (use) $X' = \mathit{drop\_dim}(X, i)$. }
\FOR {each stmt and dimension $X[...i...] = ...$}
\STATE {if $i$ is a dead dimension on exit from stmt $X[...i...] = ...$, remove $i$ from $X$ (all defs and uses)}
%\COMMENT{ A stmt v[...i] is a "dead dimension i" if all edges out of FOR i = 0; ... loop end at drop\_dim(v, i)}
\ENDFOR

\COMMENT { Now clean up drop\_dim and raise\_dim }
\FOR {each $X' =  \mathit{drop\_dim(X,i)}$ }
\STATE {replace with $X' = X$ if $i$ is dead in $X$.}
\ENDFOR
\STATE {do (1) (extended) constant propagation, (2) copy propagation and (3) dead code elimination to get rid of redundant variables and raise and drop dimension statements}

@@

\subsection{Correctness Argument}

@@

\subsection{Extension with Array Writes}
\label{sec:extension_with_writes}

>>>>

%\ana{Need an intro/overview here!}
%\ana{Need to define notion of canonical access!}
%\ana{Talk about MPC Source, opportunities for vectorization, algorithm unique to MPC Source.}

This section describes our vectorization algorithm. While vectorization is a longstanding problem, 
and we build upon existing work on scalar expansion and classical loop vectorization~\cite{Allen:1987}, 
our algorithm is unique as it works on the MPC Source SSA-form representation. We posit that vectorization 
over MPC Source is a new problem that warrants a new look, in part because of MPC's unique linear 
structure and in part because vectorization meshes in with other MPC-specific optimizations in non-trivial ways
(other works have explored manual vectorization and protocol mixing in an ad-hoc way, e.g.,~\cite{NDSS:DemSchZoh15,CCS:BDKKS18,Ishaq:2019}).

\secref{sec:dependence} describes our dependence analysis and ~\secref{sec:expansion} describes scalar expansion,
which lifts scalars (and arrays) to the corresponding loop dimensionality to create opportunities for vectorization. 
\secref{sec:basic_vectorization} describes our core vectorization algorithm and~\secref{sec:correctness} argues correctness of the vectorization transformation.
\secref{sec:array_writes} extends vectorization with array writes.

@@

We build a dependence graph where the nodes are the MPC Source statements and the edges represent the def-use relations.

\paragraph{Def-use Edges} We distinguish the following def-use edges:

@@

\paragraph{Closures}
We define $\mathit{closure}(n)$ where $n$ is a PHI-node. Intuitively, it computes the set of nodes (i.e., statements) that form a dependence cycle with $n$. The closure of $n$ is defined as follows:
\begin{itemize}
\item $n$ is in $\mathit{closure}(n)$
\item $X$ is in $\mathit{closure}(n)$ if there is a same-level path from $n$ to $X$, and $X \rightarrow n$ is a same-level back-edge.
\item $Y$ is in $\mathit{closure}(n)$ if there is a same-level path from $n$ to $Y$ and there is a same-level path from $Y$ to some $X$ in $\mathit{closure}(n)$.
\end{itemize}

\subsection{Scalar Expansion}
\label{sec:expansion}

@@

An important component of our algorithm is the scalar expansion to the corresponding loop dimensionality, 
which is necessary to expose opportunities for vectorization.
In the Biometric example, {\sf d = S[i*D+j] - C[j]} equiv. to {\sf d = S[i,j] - C[j]}, which gave rise to $N*D$ subtraction operations in the sequential schedule,
is lifted. The argument arrays {\sf S} and {\sf C} are lifted and the scalar {\sf d} is lifted: {\sf d[i,j] = S[i,j] - C[i,j]}.
The algorithm then detects that the statement can be vectorized.

\paragraph{Arrays} Conceptually, we treat all variables as arrays. There are three kinds of arrays. 
%\ana{Arrays is not a good name here.}

\begin{itemize}
\item Scalars: These are scalar variables we expand into arrays for the purposes of vectorization.
For those, all writes are canonical writes and all reads are canonical reads. We will \emph{raise dimension} when
a scalar gives rise to an outer-to-inner dependence edge (e.g., {\sf sum!2} in line 6 of the MPC source code will be raised to a 1-dimensional
array since {\sf sum!2} is used in the inner $j$-loop). We will \emph{drop dimension} when a scalar
gives rise to an inner-to-outer dependence edge (e.g., {\sf sum!2} for which the lifted inner loop computes
{\sf D} values, but the outer loop only needs the last one.)

@@

%In the sections below we detail the $\mathit{raise\_dim}$ (raise dimension) and $\mathit{drop\_dim}$ (drop dimension) operations,
%followed by our vectorization algorithm.

@@

\paragraph{Raise dimension} 
The $\mathit{raise\_dim}$ function expands a scalar (or array). 
There are two conceptual versions of $\mathit{raise\_dim}$. One reshapes an arbitrary access into a canonical read access in the corresponding loop. 
%The signature of $\mathit{raise\_dim}$ is as follows. 
It takes the original array, the access pattern function $f(i,j,k)$ in loop nest $i,j,k$ and the loop bounds
$((i:I),(j:J),(k:K))$:
\[ \mathit{raise\_dim}(A, f(i,j,k), ((i:I),(j:J),(k:K))) \]
It produces a new 3-dimensional array ${\sf A'}$ by iterating over $i,j,k$ and setting each element of  ${\sf A'}$ as follows:
\[ {\sf A'}[i,j,k] =  {\sf A}[f(i,j,k)] \]
The end result is that uses of ${\sf A}[f(i,j,k)]$ in loop nest $i,j,k$ are replaced with canonical read-accesses to ${\sf A'}[i,j,k]$
that can be vectorized. In the running Biometric example, ${\sf C'} = \mathit{raise\_dim}({\sf C}, j, (i:N,j:D))$ lifts the
1-dimensional array {\sf C} into a 2-dimensional array. The $i,j$ loop now accesses ${\sf C'}$ in the canonical way, ${\sf C'}[i,j]$.
Similarly, ${\sf S'} = \mathit{raise\_dim}({\sf S}, i*D+j, (i:N,j:D))$ tries to lift {\sf S}, but the operation turns into a no-op
because {\sf S} is already a 2-dimensional array and the read access is canonical.

The other version of $\mathit{raise\_dim}$ lifts a lower-dimension array into a
higher-dimension for access in a nested loop. It is necessary when processing outer-to-inner dependences. 
Here {\sf A} is an $i$-array and raise dimension adds two additional dimensions:
\[ \mathit{raise\_dim}({\sf A}, (j:J,k:K)) \]
This version is reduced to the above version by adding the access pattern function, which is just $i$:
\[ \mathit{raise\_dim}({\sf A}, i, (j:J,k:K)) \]

\paragraph{Drop dimension}  
The corresponding $\mathit{drop\_dim}$ is carried out when an array written in an inner loop is used in an enclosing loop.
It takes a higher dimensional array, say $i,j,k$ and removes trailing dimensions, say $j,k$:
\[ \mathit{drop\_dim}({\sf A}, (j:J,k:K)) \]
It iterates over $i$ and takes the result at the maximal index of $j$ and $k$, i.e., the result at the last iterations of $j$ and $k$:
\[ {\sf A'}[i] = A[i,J-1,K-1] \]

\subsection{Basic Vectorization}
\label{sec:basic_vectorization}

We present our algorithm followed by an example on Biometric.

@@

\FOR {each MPC $\mathit{stmt}: \x = \mathit{Op}(\y_1,\y_2)$ in loop $i,j,k$}
%\FOR {each $Y_n$}
\FOR {each argument $\y_n$} %def-use edge $\mathit{stmt}' (\mbox{def of } Y) \rightarrow \mathit{stmt} (\mbox{def of } X)$}
\STATE {case def-use edge $\mathit{stmt}' (\mbox{def of } \y_n) \rightarrow \mathit{stmt} (\mbox{def of } \x)$ of} \\
\STATE \hspace{0.25cm} {same-level: $\y'_n$ is $\y_n$} \\
\STATE \hspace{0.25cm} {outer-to-inner: add $\y'_n[i,j,k] = \mathit{raise\_dim}(\y_n)$ at $\mathit{stmt'}$}
\STATE \hspace{0.25cm} {(more precisely, right after $\mathit{stmt'}$)}
\STATE \hspace{0.25cm} {inner-to-outer: add $\y'_n[i,j,k] = \mathit{drop\_dim}(\y_n)$ at $\mathit{stmt}$}
\STATE \hspace{0.25cm} {(more precisely, in loop of $\mathit{stmt}$ right after loop of $\mathit{stmt'}$)}
\ENDFOR

\COMMENT { Optimistically vectorize all. $I$ means vectorized dimension. }
%\FOR {each MPC $\mathit{stmt}: X = \mathit{Op}(Y_1,Y_2)$ in loop $i,j,k$}
\STATE {change to $\x[I,J,K] = \mathit{Op}(\y'_1[I,J,K],\y'_2[I,J,K])$ }
\ENDFOR

\COMMENT { Phase 2: Recreating for-loops for cycles; vectorizable stmts hoisted up. }

\FOR {each dimension $d$ from highest to 0}
\FOR {each PHI-node $n$ in loop $i_1,...,i_d$}
\STATE {compute $\mathit{closure}(n)$}
\ENDFOR

@@

\WHILE {there are closure $cl_1$ and $cl_2$ that intersect}
\STATE {merge $cl_1$ and $cl_2$}
\ENDWHILE
\FOR {each closure $cl$ (after merge)}
\STATE {create \python{for} $i_d$ \python{in} ... loop}
\STATE {add PHI-nodes in $cl$ to header block}
\STATE {{\color{red}{add target-less PHI-node for A if $cl$ updates array A}}}
\STATE {add statements in $cl$ to loop in some order of dependences}
\STATE \COMMENT { Dimension is not vectorizable: }
\STATE {change ${I_d}$ to $i_d$ in all statements in loop}
\STATE {treat \python{for}-loop as monolith node for def-uses: e.g., some def-use edges become same-level.}
\ENDFOR
{
\color{red}
\FOR {each target-less PHI-node {\sf A!1 = PHI(A!0,A!k)}}
\STATE {in vectorizable stmts, replace use of {\sf A!1} with {\sf A!0}}
\STATE {discard PHI-node if not used in any $cl$, replacing {\sf A!1} with {\sf A!0} or {\sf A!k} appropriately}
\ENDFOR
}
\ENDFOR

@@

\COMMENT { A dimension $i$ is dead on exit from stmt $\x[...i...] = ...$ if all def-uses with targets outside of the enclosing
\python{for} $i$ ... MOTION loop end at target (use) $\x' = \mathit{drop\_dim}(\x, i)$. }
\FOR {each stmt and dimension $\x[...i...] = ...$}
\STATE {if $i$ is a dead dimension on exit from stmt $\x[...i...] = ...$, remove $i$ from $\x$ (all defs and uses)}
%\COMMENT{ A stmt v[...i] is a "dead dimension i" if all edges out of FOR i = 0; ... loop end at drop\_dim(v, i)}
\ENDFOR

\COMMENT { Now clean up drop\_dim and raise\_dim }
\FOR {each $\x' =  \mathit{drop\_dim(\x,i)}$ }
\STATE {replace with $\x' = \x$ if $i$ is dead in $\x$.}
\ENDFOR
\STATE {do (1) (extended) constant propagation, (2) copy propagation and (3) dead code elimination to get rid of redundant variables and raise and drop dimension statements}

@@

\subsection{Correctness Argument}
\label{sec:correctness}

@@

\subsection{Extension with Array Writes}
\label{sec:array_writes}
